---
title: "EDS 222: Final Project"
author: "Carly Caswell"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load all the packages needed here
library(tidyverse)
library(readr)
library(gt)
library(tufte)

# Set your filepath here! Or, set this up as an .Rproj if you'd like.
rootdir <- ("~/Documents/MEDS/Fall_Q2/EDS-222-Stats/Assignments")
datadir <- file.path(rootdir,"eds222-final","data")
```

ABOUT:

I am interested in the relationship between and .

DATA:

Dataset 1 - Titled "xyz" dataset called, this contains x,y, z and comes from SOURCE. I will primarily be using variables a,b,c, which contain ...... "the mean number of days with minimum temperature below freezing (mean over the years 1931-1960)."

Dataset 2 - Titled "xyz" dataset from COUNTRY. These data were obtained from the [ERA5 database](https://www.ecmwf.int/en/forecasts/datasets/reanalysis-datasets/era5), a product made available by the European Centre for Medium-Range Weather Forecast. The high-resolution hourly gridded data were aggregated to the municipality by month level -- that is, each observation in these data report a monthly average temperature value and a monthly cumulative precipitation value for one of the 1,123 municipalities across the country.[\^3]

```{r}
# Reading in the data:
    

#Making them dataframes:
 = data.frame()
 = data.frame()


```

EXPLORATION:

```{r}
#Let's take a look at some of this data further:
head()


#Creating a scatter plot to view the relationship:
    ggplot(data=, aes(x=, y = )) + 
      geom_point() + theme_minimal() 
      

#Computing the mean and standard dev for the Frost days of each region:
days_below_zero = state_frost %>% 
  group_by(state.region) %>% 
  summarize("Frost Mean" = mean(Frost), "Frost SD" = sd(Frost)) 

#Reviewing the results table:
print(days_below_zero)
#The region with the highest variance in number of frost days is the West region
```

**I see a mostly positive/negative relationship between y in units and x. The correlation seems to be mostly strong with a few outliers that could have an impact on the overall correlation coefficient.**

**I think this relationship makes intuitive sense because the higher the temperature the drier and more fire-prone conditions there will be.**

MY QUESTION:

Is the mean number of frost days different in the North Central region than in the South?

My null and alternative hypotheses:

$$H_{O}: \mu_{north} - \mu_{south} = 0$$

$$H_{A}: \mu_{north} - \mu_{south} \neq 0$$

ANALYSIS:

```{r}
#Running a regression:
summary(lm(area ~ temp, data =ffires))

#Running a t.test on the filtered data:
t.test(Frost ~ state.region, data = filtered_state_frost, conf.level = 0.95)

#Construct a 95% confidence interval for your slope coefficient:
conf_interval <- confint(murders, level = .95)[2, ] #using the confint function

print(paste('I am 95% confident the range', round(conf_interval[1], 3), 'and',  round(conf_interval[2], 3), 'contains the true population difference between murder rates per 100,000 people and average number of frost days between 1931 and 1960.'))

```

Interpreting the intercept and slope coefficients:

\*\*When the average number of x is 0, the b1 is 11.376 per 100,000 people between the years 1931-1960.

**For every 1 day increase in average frost days per year, the murder rate per 100,000 people decreases by -.03827.**

Is there a statistically significant relationship between frost days on murder rates? At what significant level is this effect significant?

**Yes, p value is .00005 with a significance level of .05 so this proves that there is a significant relationship between Frost days and murder rates**

**The p-value is just slightly different than the manual calculation, but I still end up with the same conclusion (reject the null). This is what I'd expect since I'm using a q normal distribution in my manual approach and a t distribution in the function approach.**
